---
title: "10-exploratory-data-analysis-EDA"
format: html
editor: 
  markdown: 
    wrap: 72
---

## 10.1 Introduction

-   exploratory data analysis (EDA) is an iterative process of
    visualizing, transforming, and modeling data to discover and
    understand the elements relevant for your goals (often modeling for
    prediction/forecasting or simulation). EDA process:

1.  generate questions about data
2.  search for answers by visualizing, transforming, and modelling data
3.  use what was learned to refine questions and/or generate new
    questions (start back at step 1)

-   data cleaning is one application of EDA: goal is to answer whether
    data quality meets expectations or not

### 10.1.1 Prerequisites

```{r}
library(tidyverse)
library(patchwork)
sessionInfo()

# https://stackoverflow.com/questions/21967254/how-to-write-a-reader-friendly-sessioninfo-to-text-file
if (!file.exists("./session_info.txt")) {
  writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
}
```

## 10.2 Questions

EDA is creative process, not exact science. Be sure to ask thoughtful
questions and learn as you go - can't be perfect from onset but can
quickly learn by following up each quality question with new quality
question(s) based on findings

Again, each dataset generally requires its own set of questions needed
to gain useful insights based on your goal, but here are two questions
that are always useful starting points.

1.  What type of variation occurs within my variables?
2.  What type of covariation occurs between my variables?

## 10.3 Variation

**Variation**: tendency of values of a variable to change from
measurement to measurement. measuring a continuous variable twice will
always yield different values given enough precision (decimal places).
two sources of variation:

1.  when observing some instance of a variable: measurement errors
    across replicates of the same observation subject.

-   variation due to instrument limitation: no instrument is infinitely
    precise
-   variation by time: replicating exact same conditions for some
    instance is nearly impossible
-   e.g. weight of a single person measured multiple times. we can't get
    infinite decimal places on their mass and, for any specific decimal,
    there's infinitely low probability that their mass is defined
    uniquely to any given decimal place e.g. given person A \[1\] does
    person A weigh 150.5 kg, 50.49 kg, 150.499 kg, etc. \[2\] what if
    they changed their diet between days or just ate before being
    weighed?

2.  when observing different instances of a variable: different
    observations/subjects of a variable, not replicates (e.g. weight of
    different people being measured)

-   *QUESTION FOR LATER* how hard is it to distinguish these two
    variation causes?

### case study begin

let's explore the diamond dataset, starting with the carat variable

-   best way to understand pattern of variation is to visualize it.
    histograms are great for continuous variables
-   twofold key to good questions, ask variations of:
    -   what do i want to know (curiosity)
    -   how can what i see be misleading (skepticism)

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

### 10.3.1 Typical values

in bar charts and histograms,

-   tall bars mean common values
-   no bars means values not in dataset

questions to ask:

-   Which values are the most common? Why?
-   Which values are rare? Why? Does that match your expectations?
-   Can you see any unusual patterns? What might explain them?

```{r}
# let's look at carat for smaller diamonds
smaller <- diamonds |> 
  filter(carat < 3)

ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

this plot begets these questions:

-   Why are there more diamonds at whole carats and common fractions of
    carats?
-   Why are there more diamonds slightly to the right of each peak than
    there are slightly to the left of each peak?

visualization can also reveal "clustering" (aka "sub grouping") of data.

-   How are the observations within each subgroup similar to each other?
-   How are the observations in separate clusters different from each
    other?
-   How can you explain or describe the clusters?
-   Why might the appearance of clusters be misleading?

### 10.3.2 Unusual values

outliers = data that doesn't follow rest of dataset's pattern

causes of outliers?

1.  data entry errors
2.  extreme values that are rare but possible
3.  rare but important discoveries

with lots of data, outliers can be hard to see

```{r}
# primary evidence of outliers in plot below is the strangely wide x axis, as 
# most of data falls in the 0 to 10 range
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
```

let's zoom in on the y axis using `coord_cartesian()`. note:
`coord_cartesian()` zooms while `xlim()` and `ylim()` which throw out
values outside the limits

```{r}
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

we see 3 unusual bins at 0, \~30, and \~60. let's extract them

```{r}
unusual <- diamonds |> 
  filter(y < 3 | y > 20) |> 
  select(price, x, y, z) |>
  arrange(y)
unusual
```

`x`, `y`, `z` are measure each of the 3 dimensions of the diamonds. 0 is
impossible, so we know these aren't real measurements! these values must
have been incorrectly recorded, and should be `NA`. EDA has revealed
missing values that searching for `NA` wouldn't have found

**what to do when you find outliers?** repeat analysis with and without
outliers. if they have minimal effect and can't figure out why they're
there, then simply omit them. if they have large effect, then must
have a valid justification for removing them.

### 10.3.3 Exercises
1. Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

- answer: x, y, and z are explored below. according to `?diamonds` x is the length, y is width, z is "depth" (different from the var `depth`). when looking at the distribution and limits on the histograms/description page, x appears to be smaller than y. then z must be that third dimension orthogonal to both. there appears to be 20 rows in diamonds that have zero values for x, y, and/or z (which is invalid). i suggest either correcting them by back calculating from `depth`, or omitting these values.

```{r}
diamonds |>
  ggplot(aes(x = x)) +
  geom_histogram(binwidth = 0.5) + 
  ggtitle("Distribution of x")

# are there extreme values of x? not really.
diamonds |>
  ggplot(aes(x = x)) +
  geom_histogram(binwidth = 0.5) + 
  coord_cartesian(ylim = c(0, 7500)) +
  ggtitle("Distribution of x")

# what do the "extreme" values of x look like in terms of their whole observation set?
diamonds |>
  filter(
    x >= 9 | x < 3.75
  )

# strange, there are some 0 measurements in x, y, and/or z above. this isnt
# possible, otherwise the diamonds would be 2D instead of 3D. they couldn't have
# a "depth" measurement that is nonzero or non-infinity.
# what do the observations with zero values look like?
diamonds |>
  filter(
    x == 0 | y == 0 | z == 0
  ) |>
  arrange(x, y, z)
```
```{r}
diamonds |>
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 0.05) + 
  ggtitle("Distribution of y")

diamonds |>
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 0.5) + 
  ggtitle("Distribution of y")

diamonds |>
  ggplot(aes(x = y)) +
  geom_histogram(binwidth = 5) + 
  ggtitle("Distribution of y")
```

```{r}
diamonds |>
  ggplot(aes(x = z)) +
  geom_histogram(binwidth = 0.05) + 
  ggtitle("Distribution of z")

diamonds |>
  ggplot(aes(x = z)) +
  geom_histogram(binwidth = 0.5) + 
  ggtitle("Distribution of z")

diamonds |>
  ggplot(aes(x = z)) +
  geom_histogram(binwidth = 5) + 
  ggtitle("Distribution of z")
```


2. Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

- answer: For a binwidth of $1000, the second smallest bin contains a large proportion of the data, 18261 diamonds. There are 53940 diamonds in the dataset, so 18621/53940 = 34.5% of the data. This bin includes diamonds in the price range [500, 1500). This is unique, but I don't think it's surprising. Most jewelers probably want to keep most cuts of diamonds in the 1k price range to incentive people to buy them, as I think people will be less willing to buy more expensive diamonds (or perhaps unable to buy more expensive). 

```{r}
for (cur_binwidth in 10^seq(0,4)) {
  print(cur_binwidth)
  tmp <- diamonds |>
    ggplot(mapping = aes(price)) +
    geom_histogram(binwidth = cur_binwidth) +
    ggtitle(paste0("Distribution of price"), subtitle = paste0("binwidth = ", cur_binwidth)) + 
    stat_bin(binwidth = cur_binwidth, 
             geom = 'text', 
             color = 'red', 
             aes(label = after_stat(count)),
             angle = 90,
           position = position_stack(vjust = 0.5))
  print(tmp)
}
```
```{r}
diamonds |>
  filter(price < 2000) |>
  arrange(price)
```

```{r}
p <- ggplot(diamonds, aes(price)) +
  geom_histogram(binwidth = 1000)

bins <- ggplot_build(p)$data[[1]]

bins
```


3. How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

- answer: There are 23 diamonds that are 0.99 carat and 1558 diamonds that are 1.00 carat. I originally suspect that the difference is due to some diamonds being rounded but not others (due to convenience and/or instrument precision for people contributing to the dataset), but I think that would meant that certain multiples of a number should have big spikes in the histogram at regular intervals and low values everywhere else. Instead of that we get this weird pattern where a certain value has a large bar then the bars gradually decrease in size until a large spike. The peaks form a repeating "exponential decay" pattern. This could be a rounding pattern, perhaps it's rare to measure certain values of carat so precisely.

```{r}
diamonds |>
  count(carat) |>
  filter(carat >= 0.95 & carat <= 1.05)
```

```{r}
set_binwidths = c(0.01, 0.05, 0.10, 0.25, 0.5, 1)
for (cur_binwidth in set_binwidths) {
  print(cur_binwidth)
  tmp <- diamonds |>
    ggplot(mapping = aes(carat)) +
    geom_histogram(binwidth = cur_binwidth) +
    ggtitle(paste0("Distribution of carat"), subtitle = paste0("binwidth = ", cur_binwidth)) 
    
    
  if (cur_binwidth > 0.05)
    tmp = tmp + 
      stat_bin(binwidth = cur_binwidth, 
             geom = 'text', 
             color = 'red', 
             aes(label = after_stat(count)),
             angle = 90,
           position = position_stack(vjust = 0.5))
  else
      tmp = tmp +
        scale_x_continuous(breaks = scales::breaks_extended(n = 50)) +
        theme(
          axis.text.x = element_text(color="black", angle=90)
        ) +
      stat_bin(binwidth = cur_binwidth, 
       geom = 'text', 
       color = 'red', 
       aes(label = after_stat(count)),
       angle = 90,
       size = 1.5,
       position = position_stack(vjust = 1)
      )
    
  print(tmp)
}
```
```{r}
test_carat_multiples <- seq(0.01,0.5,0.01)

for (cur_carat_multiple in test_carat_multiples) {
  tmp <- diamonds |>
    filter(
      ((carat*100) %% (cur_carat_multiple*100)) == 0
    )
  
  # print(tmp)
  print(paste0("cur_carat_multiple = ", cur_carat_multiple))
  print(paste0("num diamonds at this carat multiple = ", nrow(tmp)))
  print("---")
}
```

4. Compare and contrast coord_cartesian() vs. xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

- answer: see comments and plots below. when setting binwidth, coord_cartesian behaves as generally desired (a zooming function) rather than removing out of bounds bars like xlim/ylim. when the binwidth is not set, it seems the default value for bins (30) forces both to appear to work. as explained in the text though, xlim/ylim will silently remove values from the dataset when doing the counting statistic, whereas coord_cartesian will count all values but may not display some depending on the bounds

setting binwidth
```{r}
tmp <- ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 1) + 
      stat_bin(binwidth = 1, 
             geom = 'text', 
             color = 'red', 
             aes(label = after_stat(count)),
             angle = 90,
           position = position_stack(vjust = 0.5))

# try to cut out some bars by focusing on smaller x axis? works with coord_cartesian!
tmp +
  coord_cartesian(xlim = c(0,2))

# try to cut bar in half by focusing on smaller y axis? works with coord_cartesian!
tmp +
  coord_cartesian(ylim = c(0,25000))

# try to cut out some bars by focusing on smaller x axis? DOESNT work with xlim! removes the bars containing out of limit bounds
tmp +
  xlim(c(0,2))

# try to cut bar in half by focusing on smaller y axis? DOESNT workswith ylim! bars that would be presented outside the y axis range are simply removed
tmp +
  ylim(c(0,25000))
```

not setting binwidth (letting default `bins = 30` determine binwidth)
```{r}
tmp <- ggplot(diamonds, aes(x = carat)) +
  geom_histogram() + 
      stat_bin( 
             geom = 'text', 
             color = 'red', 
             aes(label = after_stat(count)),
             angle = 90,
           position = position_stack(vjust = 0.5))

# try to cut out some bars by focusing on smaller x axis? works with coord_cartesian!
tmp +
  coord_cartesian(xlim = c(0,2))

# try to cut bar in half by focusing on smaller y axis? works with coord_cartesian!
tmp +
  coord_cartesian(ylim = c(0,25000))

# try to cut out some bars by focusing on smaller x axis? DOESNT work with xlim! removes the bars containing out of limit bounds
tmp +
  xlim(c(0,2))

# try to cut bar in half by focusing on smaller y axis? DOESNT workswith ylim! bars that would be presented outside the y axis range are simply removed
tmp +
  ylim(c(0,25000))
```

## 10.4